### 2022-08-05 chapter1/chapter.tex
ch 7/8 comments

\section{Some semigroup theory}
The evolution of Markov processes can be described by an \emph{operator semigroup}. Semigroups also arise in other contexts related to the analysis of fluid queues and fluid-fluid queues (see, for example, Sections~\ref{sec: ffq intro} and~\ref{sec: transient ffq intro}). Semigroups (and therefore Markov processes) can be characterised by their \emph{infinitesimal generator}, or generator for short. One of the aims of this thesis is to approximate the generator of a fluid queue. Here we very briefly introduce semigroups and their \emph{infinitesimal generators}.

\subsubsection{Operators}
Let \(L\) be a Banach space (a vector space with a metric and where Cauchy sequences converge to a point within the space). An operator is a mapping which takes elements of a vector space and sends them to elements of a vector space (not necessarily the same space). That is, let \(V\) and \(W\) be normed vector spaces, then \(\mathbb B:\mathcal D(\mathbb B)\subset V\to W\) is an operator. The set \(\mathcal D(\mathbb B)\) is the domain of the operator, and is not necessarily all of \(V\). When specification of the domain is unnecessary we sometimes write \(\mathbb B:V\to W\). Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a \emph{linear} operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms attached to \(V\) and \(W\), respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(M<\infty, M\in \mathbb R\), for every \(v\in V\).

@@

\subsubsection{Semigroups}  A family of linear operators is a collection of indexed linear operators \(\{\mathbb B(i)\}_{i\in I}\) where \(I\) is some index set. That is, for each \(i\in I\), \(\mathbb B(i)\) is a linear operator. A family of operators \(\{\mathbb T(t)\}_{t\geq 0}\), \(\mathbb T(t):L\to L\), is said to have the \emph{semigroup} property if \(\mathbb T(t+s)=\mathbb T(t)\mathbb T(s)\). If \(\mathbb T(0)=I\), the identity operator, then the family \(\{\mathbb T(t)_{t\geq 0}\) is known as a semigroup. If \(\mathbb T(t)f\to f\) as \(t\to 0^+\), then \(\{\mathbb T(t)\}_{t\geq 0}\) is said to be strongly continuous. If \(||\mathbb T(t)||\leq 1\) for all \(t\geq 0\) then \(\{\mathbb T(t)\}_{t\geq 0}\) is known as a contraction semigroup.

@@

% \section{Markov processes and semigroups}
% Semigroups arise naturally in Markov processes. Let \(\{X(t):t\geq 0\}\) be a Markov process with state space \(\mathcal S\). The Markov property states that for \(s,t\geq 0\), \(X(t+s)\) is independent of \(X(u),\, 0\leq u<t\) given \(X(t)\). That is, \(P(X(t+s)\in\mathcal{E} \mid X(t), X(u), 0\leq u<t)=P(X(t+s)\in\mathcal{E} \mid X(t))\). If \(P(X(t+s)\in\mathcal{E} \mid X(t))=P(X(s)\in\mathcal{E} \mid X(0))\), that is the distributions are invariant under translation by \(t\), then \(X(t)\) is said to be time-homogeneous. Where necessary, we will use the notation \(X_\mu(t)\) to denote the Markov process with initial distribution \(\mu\), i.e.~\(P(X(0)\in\cdot)=\mu(\cdot)\), and \(X_x(t)\) to be the Markov process with initial distribution \(\delta(x)\). Markov processes can be characterised by \emph{probability kernels}.

% Let \((U,\mathcal U)\) and \((V,\mathcal V)\) be two measure spaces, then a probability kernel \(\nu\), is a mapping \(\nu:U\times \mathcal V\to \mathbb R_+\) such that \(\nu(u,\mathcal{E})\) is \(\mathcal U\)-measurable in \(u\in U\) for fixed \(\mathcal{E}\in \mathcal V\), and a probability measure in \(\mathcal{E}\in \mathcal V\) for fixed \(u\in U\). We say that \(\nu\) is a probability kernel from \(U\) to \(V\).

@@

% Let \(\mu_{t}(\cdot)=P(X(t)\in \cdot)\) for \(t\geq 0\) be a one-parameter family of measures on \((\mathcal S,\mathcal F)\), and \(\nu_{t}(x,\cdot)=P(X(t)\in\cdot \mid X(0)=x)\) for \(t\geq 0\) be a one-parameter family of probability kernels from \(\mathcal S\) to \(\mathcal S\). Then \(\mu_t\) and \(\nu_{t}\) characterise the time-homogeneous Markov process \(X(t)\). The probability kernel \(\nu_t\) is referred to as a transition kernel. The kernels \(\nu_t\) form a semigroup since, using the law of total probability and time-homogeneity, 
% \begin{align*}
% \nu_{t+s}(x,\mathcal{E})&=P(X(t+s)\in\mathcal{E}\mid X(0)=x)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(t+s)\in\mathcal{E}\mid X(t)=y)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(s)\in\mathcal{E}\mid X(0)=y)
% \\&=\int_{y\in\mathcal S}\nu_t(x,\wrt y)\nu_s(y,\mathcal{E})=\nu_t\nu_s(x,\mathcal{E}).
% \end{align*}

@@

\subsection{Matrix exponential distributions}
Here we recount some facts about matrix exponential distributions. See \cite{MEinAP} for a more detailed exposition. A random variable, \(Z\), is said to have a matrix exponential distribution if it has a distribution function of the form \(1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s\), where \(\bs \alpha\) is a \(1\times p\) \emph{initial vector}, \(\bs S\) a \(p\times p\) matrix, and \(\bs s\) a \(p\times 1\) \emph{closing vector}, and \(\displaystyle e^{\bs S x} := \sum_{n=0}^\infty \cfrac{\left(\bs Sx\right)^n}{n!}\) is the matrix exponential. The density function of \(Z\) is given by \(f_Z(x) = \bs\alpha e^{\bs S x}\bs s\). The only restrictions on the parameters \((\bs \alpha, \bs S, \bs s)\) are that \(\bs \alpha e^{\bs S x} \bs s\) be a valid density function, i.e.~\(\bs \alpha e^{\bs S x} \bs s\geq 0,\) for all \(x\geq 0\) and \(\lim_{x\to\infty} 1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s = 1.\) In general, in \cite{MEinAP}, there is the possibility of an \emph{atom} (a point mass) at 0, but here we do not consider this possibility. The condition that \(\bs \alpha e^{\bs Sx}\bs s\) be a valid density imposes some properties on representations \((\bs \alpha, \bs S, \bs s)\). However, in general there is no way to determine whether, a given triplet \((\bs \alpha, \bs S, \bs s)\) is a representation of a matrix exponential distribution, or not. Nonetheless, some properties of a triplet \((\bs \alpha, \bs S, \bs s)\) are known, such as the following, which is used in the characterisation of QBD-RAPs. 
\begin{thm}[Theorem 4.1.3, \cite{MEinAP}]
	The density function of a matrix exponential distribution with representation \((\bs\alpha, \bs S, \bs s)\) can be expressed in terms of real-valued constants as 
	\begin{align}
		\psi(x)&=\sum_{j=1}^{m_1} \sum_{k=1}^{p_j} c_{jk} \cfrac{x^{k-1}}{(k-1)!} e^{\mu_jx} + \sum_{j=1}^{m+2}\sum_{k=1}^{q_j} d_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\cos(\sigma_jx) \nonumber 
		\\&\quad{}+ \sum_{j=1}^{m_2}\sum_{k=1}^{q_j}e_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\sin(\sigma_jx), \label{eqn: spec}
	\end{align}
	for integers \(m_1,\,m_2,\,p_j,\) and \(q_j\) and some real constants \(c_{jk},\,d_{jk},\,e_{jk},\,\mu_j,\,\eta_j,\) and \(\sigma_j\). Here \(\mu_j,\, j=1,\dots,m_1\) are the real eigenvalues of \(\bs S\), while \(\eta_j+i\sigma_j, \, \eta_j-i\sigma_j,\, j=1,\dots,m_2\) denote its complex eigenvalues, which come in conjugate pairs. Thus, \(m_1+2m_2\) is the total number of eigenvalues, while the dimension of the representation is given by \(\displaystyle p=\sum_{j=1}^{m+1}p_j + 2\sum_{j=1}^{m_2}q_j\). 
\end{thm}
\begin{thm}[Theorem 4.1.4, \cite{MEinAP}]\label{thm: 4.1.4}
	Consider the nonvanishing terms of the matrix exponential density (\ref{eqn: spec}), \emph{i.e.}, the terms for which \(c_{jk}\neq 0,\,d_{jk}\neq 0,\) or \(e_{jk}\neq 0\). Among the corresponding eigenvalues \(\lambda_j\), there is a real dominating eigenvalue \(\kappa\), say. That is, \(\kappa\) is real, \(\kappa \geq Re(\lambda_j)\) for all \(j\), and the multiplicity of \(\kappa\) is at least the multiplicity of every other eigenvalue with real part \(\kappa\).
\end{thm}
\begin{cor}[Corollary 4.1.5, \cite{MEinAP}]
If \((\bs \alpha, \bs S, \bs s)\) is a representation for a matrix exponential distribution, then \(\bs S\) has a real dominating eigenvalues.
\end{cor}
\begin{thm}[Theorem 4.1.6, \cite{MEinAP}]
	Let \(Z\) be a matrix-exponentially distributed random variable with density (\ref{eqn: spec}). Then the dominant real eigenvalue \(\kappa\) of Theorem~\ref{thm: 4.1.4} is strictly negative. 
\end{thm}
We define \(dev(\bs S)\) to be the real dominating eigenvalue of \(\bs S\), that is \(dev(\bs S)=\kappa\) in Theorem~\ref{thm: 4.1.4}.

@@

@@

>>>>

\section{Some semigroup theory}
The evolution of Markov processes can be described by an \emph{operator semigroup}. Semigroups also arise in other contexts related to the analysis of fluid queues and fluid-fluid queues (see, for example, Sections~\ref{sec: ffq intro} and~\ref{sec: transient ffq intro}). Semigroups (and therefore Markov processes) can be characterised by their \emph{infinitesimal generator}, or generator for short. One of the aims of this thesis is to approximate the generator of a fluid queue. Here we very briefly introduce semigroups and their \emph{infinitesimal generators}. We refer to the reader to \cite{ethierkurtz} for more details (see also \cite{kallenberg}, or the more approachable course notes of \cite{shalizi}).

\subsubsection{Operators}
Let \(L\) be a \emph{Banach space}. A Banach space is a \emph{complete, normed} vector space. Complete means that every Cauchy sequence of points in \(L\) has a limit which also lies in \(L\). Normed means that a norm is defined on \(L\). Intuitively, a norm is a function which tells us the distance between two points in \(L\).

An \emph{operator} is a mapping which takes elements of a vector space and sends them to elements of a vector space (not necessarily the same space). That is, let \(V\) and \(W\) be normed vector spaces, then \(\mathbb B:\mathcal D(\mathbb B)\to W\) is an operator. The set \(\mathcal D(\mathbb B)\subseteq V\) is the domain of the operator, and is not necessarily all of \(V\). When specification of the domain is unnecessary we sometimes write \(\mathbb B:V\to W\).

Let \(c\in \mathbb R\) and \(v\in V\), \(w\in W\). Then \(\mathbb B\) is a \emph{linear} operator if \(\mathbb B(cv) = c(\mathbb Bv)\) and \(\mathbb B(v+w)=\mathbb Bv+\mathbb Bw\). We define the norm of an operator as \(||\mathbb B||=\sup_{v\in V}\cfrac{||\mathbb Bv||_W}{||v||_V}\), where \(||\cdot||_V\) and \(||\cdot||_W\) are the norms defined on \(V\) and \(W\), respectively.  An operator is said to be bounded if \(||\mathbb B||<M\) for some \(M<\infty, M\in \mathbb R\), for every \(v\in V\).

@@

\subsubsection{Semigroups}  A family of linear operators is a collection of indexed linear operators \(\{\mathbb B(i)\}_{i\in I}\) where \(I\) is some index set. For each \(i\in I\), \(\mathbb B(i)\) is a linear operator. A family of operators \(\{\mathbb T(t)\}_{t\geq 0}\), \(\mathbb T(t):L\to L\), is said to have the \emph{semigroup} property if \(\mathbb T(t+s)=\mathbb T(t)\mathbb T(s)\). If \(\mathbb T(0)=I\), the identity operator, then the family \(\{\mathbb T(t)_{t\geq 0}\) is known as an operator semigroup. If \(f\in L,\, \mathbb T(t)f\to f\) as \(t\to 0^+\), then \(\{\mathbb T(t)\}_{t\geq 0}\) is said to be strongly continuous. If \(||\mathbb T(t)||\leq 1\) for all \(t\geq 0\) then \(\{\mathbb T(t)\}_{t\geq 0}\) is known as a contraction semigroup.

@@

\subsubsection{Markov processes and semigroups}
Semigroups arise naturally in Markov processes. Let \(\{X(t):t\geq 0\}\) be a Markov process with state space \(\mathcal S\). The Markov property states that, given \(X(t)\), then for \(s,t\geq 0\), \(X(t+s)\) is independent of \(X(u),\, 0\leq u<t\). That is, \[P(X(t+s)\in\mathcal{E} \mid X(t), X(u), 0\leq u<t)=P(X(t+s)\in\mathcal{E} \mid X(t)).\] 
If \(P(X(t+s)\in\mathcal{E} \mid X(t))=P(X(s)\in\mathcal{E} \mid X(0))\), are invariant under translation by \(t\), then \(X(t)\) is said to be time-homogeneous.

% Markov processes can be characterised by \emph{probability kernels}. Let \((U,\mathcal U)\) and \((V,\mathcal V)\) be two measure spaces, then a probability kernel \(\nu\), is a mapping \(\nu:U\times \mathcal V\to \mathbb R_+\) such that \(\nu(u,\mathcal{E})\) is \(\mathcal U\)-measurable in \(u\in U\) for fixed \(\mathcal{E}\in \mathcal V\), and a probability measure in \(\mathcal{E}\in \mathcal V\) for fixed \(u\in U\). We say that \(\nu\) is a probability kernel from \(U\) to \(V\).

@@

Consider the expectation
\[\mathbb E[f(X(t))\mid X(0)=x]=\int_{y\in\calS} f(y) P(X(t)\in\wrt y\mid X(0)=x),\]
where \(f\) is some bounded function. Expectation is an operator which is acting on a function \(f\). The expectation depends on \(t\) and \(x\) as these define the distribution with which we are taking the expectation, hence, let us write 
\[\mathbb T(t)f(x) = \mathbb E[f(X(t))\mid X(0)=x].\]
The result of the operator acting on \(f\), \(\mathbb T(t)f(x)\), is another function.

By the Markov property, for some \(s\in[0,t]\), 
\begin{align*}
	&\int_{y\in\calS} f(y) P(X(t)\in\wrt y \mid X(0)=x)
	\\&=\int_{y\in\calS}\int_{z\in\mathcal S}f(y) P(X(t)\in\wrt y\mid X(s)=z) P(X(s)\in\wrt z\mid X(0)=x)
	\\&=\int_{y\in\calS}\int_{z\in\mathcal S}\mathbb E[f(X(t))\mid X(s)=z] P(X(s)\in\wrt z\mid X(0)=x)
	\\&=\int_{y\in\calS}\int_{z\in\mathcal S} \mathbb E[f(X(t-s))\mid X(0)=z]P(X(s)\in\wrt z\mid X(0)=x),
	\\&=\int_{y\in\calS}\int_{z\in\mathcal S} (\mathbb T(t-s)f(z))P(X(s)\in\wrt z\mid X(0)=x),
	\\&=\mathbb T(s) (\mathbb T(t-s)f)(x),
\end{align*}
where the third equality holds by time-homogeneity. Hence, \(\{\mathbb T(t)(\cdot)\}_{t\geq 0}\) is an operator semigroup. Moreover, since \(f\) is bounded, then \(\mathbb T(t)f(x) = \mathbb E[f(X(t))\mid X(0)=x] \leq F\), which means that \(\{\mathbb T(t)\}\) is a contraction semigroup.

% Let \(\nu_{t}(x,\cdot)=P(X(t)\in\cdot \mid X(0)=x)\) for \(t\geq 0\) be a one-parameter family of probability kernels from \(\mathcal S\) to \(\mathcal S\). Then \(\mu_t\) and \(\nu_{t}\) characterise the time-homogeneous Markov process \(X(t)\). The probability kernel \(\nu_t\) is referred to as a transition kernel. The kernels \(\nu_t\) form a semigroup since, using the law of total probability and time-homogeneity, 
% \begin{align*}
% \nu_{t+s}(x,\mathcal{E})&=P(X(t+s)\in\mathcal{E}\mid X(0)=x)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(t+s)\in\mathcal{E}\mid X(t)=y)
% \\&=\int_{y\in\mathcal S}P(X(t)\in\wrt y\mid X(0)=x)P(X(s)\in\mathcal{E}\mid X(0)=y)
% \\&=\int_{y\in\mathcal S}\nu_t(x,\wrt y)\nu_s(y,\mathcal{E})=\nu_t\nu_s(x,\mathcal{E}).
% \end{align*}

@@

\subsection{Matrix exponential distributions}
Here we recount some facts about matrix exponential distributions. See \cite{MEinAP} for a more detailed exposition. A random variable, \(Z\), is said to have a matrix exponential distribution if it has a distribution function of the form \(1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s\), where \(\bs \alpha\) is a \(1\times p\) \emph{initial vector}, \(\bs S\) a \(p\times p\) matrix, and \(\bs s\) a \(p\times 1\) \emph{closing vector}, and \(\displaystyle e^{\bs S x} := \sum_{n=0}^\infty \cfrac{\left(\bs Sx\right)^n}{n!}\) is the matrix exponential. The density function of \(Z\) is given by \(f_Z(x) = \bs\alpha e^{\bs S x}\bs s\). The only restrictions on the parameters \((\bs \alpha, \bs S, \bs s)\) are that \(\bs \alpha e^{\bs S x} \bs s\) be a valid density function, i.e.~\(\bs \alpha e^{\bs S x} \bs s\geq 0,\) for all \(x\geq 0\) and \(\lim_{x\to\infty} 1-\bs \alpha e^{\bs Sx}(-\bs S)^{-1}\bs s = 1.\) In general, in \cite{MEinAP}, there is the possibility of an \emph{atom} (a point mass) at 0, but here we do not consider this possibility. The condition that \(\bs \alpha e^{\bs Sx}\bs s\) be a valid density imposes some properties on representations \((\bs \alpha, \bs S, \bs s)\). However, in general there is no way to determine whether, a given triplet \((\bs \alpha, \bs S, \bs s)\) is a representation of a matrix exponential distribution, or not. Nonetheless, some properties of a triplet \((\bs \alpha, \bs S, \bs s)\) are known, such as the following, which is used in the characterisation of QBD-RAPs. 
\begin{thm}[Theorem 4.1.3, \cite{MEinAP}]
	The density function of a matrix exponential distribution with representation \((\bs\alpha, \bs S, \bs s)\) can be expressed in terms of real-valued constants as 
	\begin{align}
		\psi(x)&=\sum_{j=1}^{m_1} \sum_{k=1}^{p_j} c_{jk} \cfrac{x^{k-1}}{(k-1)!} e^{\mu_jx} + \sum_{j=1}^{m+2}\sum_{k=1}^{q_j} d_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\cos(\sigma_jx) \nonumber 
		\\&\quad{}+ \sum_{j=1}^{m_2}\sum_{k=1}^{q_j}e_{jk}\cfrac{x^{k-1}}{(k-1)!}e^{\eta_jx}\sin(\sigma_jx), \label{eqn: spec}
	\end{align}
	for integers \(m_1,\,m_2,\,p_j,\) and \(q_j\) and some real constants \(c_{jk},\,d_{jk},\,e_{jk},\,\mu_j,\,\eta_j,\) and \(\sigma_j\). Here \(\mu_j,\, j=1,\dots,m_1\) are the real eigenvalues of \(\bs S\), while \(\eta_j+i\sigma_j, \, \eta_j-i\sigma_j,\, j=1,\dots,m_2\) denote its complex eigenvalues, which come in conjugate pairs. Thus, \(m_1+2m_2\) is the total number of eigenvalues, while the dimension of the representation is given by \(\displaystyle p=\sum_{j=1}^{m+1}p_j + 2\sum_{j=1}^{m_2}q_j\). 
\end{thm}
\begin{thm}[Theorem 4.1.4, \cite{MEinAP}]\label{thm: 4.1.4}
	Consider the non-vanishing terms of the matrix exponential density (\ref{eqn: spec}), \emph{i.e.}, the terms for which \(c_{jk}\neq 0,\,d_{jk}\neq 0,\) or \(e_{jk}\neq 0\). Among the corresponding eigenvalues \(\lambda_j\), there is a real dominating eigenvalue \(\kappa\), say. That is, \(\kappa\) is real, \(\kappa \geq Re(\lambda_j)\) for all \(j\), and the multiplicity of \(\kappa\) is at least the multiplicity of every other eigenvalue with real part \(\kappa\).
\end{thm}
\begin{cor}[Corollary 4.1.5, \cite{MEinAP}]
If \((\bs \alpha, \bs S, \bs s)\) is a representation for a matrix exponential distribution, then \(\bs S\) has a real dominating eigenvalues.
\end{cor}
\begin{thm}[Theorem 4.1.6, \cite{MEinAP}]
	Let \(Z\) be a matrix-exponentially distributed random variable with density (\ref{eqn: spec}). Then the dominant real eigenvalue \(\kappa\) of Theorem~\ref{thm: 4.1.4} is strictly negative. 
\end{thm}
We define \(dev(\bs S)\) to be the real dominating eigenvalue of \(\bs S\), that is \(dev(\bs S)=\kappa\) in Theorem~\ref{thm: 4.1.4}.

@@

\subsubsection*{Concentrated matrix exponential distributions}
Recently, the class of \emph{concentrated matrix exponential distributions} (CMEs) has been investigated \citep{ert2006,hstz2016,ert2006,hht2020,mt2021}. As the name suggests, concentrated matrix exponential distributions are matrix exponential distributions with a very low coefficient of variation (variance relative to the mean). As the order, \(p\), of the representation increases, the variance of concentrated matrix exponential distributions decreases at rate approximately \(\mathcal O(1/p^2)\). For comparison, the variance of an Erlang distribution, the most concentrated Phase-type distribution, decreases at rate \(\mathcal O(1/p)\) as the order \(p\) increases \citep{as1987}. In this thesis we use CMEs to approximate deterministic events. For a given order, concentrated matrix exponential distributions have a much lower coefficient of variation than any phase type of the same order, and therefore a better ability to model determinism.

The class of concentrated matrix exponential distributions (CMEs) found numerically in \citep{hht2020} and computationally efficient expressions for the density and moments of CMEs are provided in \citep{hht2020}.

CMEs exist for all orders, however, in our computations we only use CMEs with odd order. The justification for considering representations of odd orders only is that the variance of CME distributions of orders \(2p\) and \(2p-1\) are relatively similar and therefore have similar abilities to represent the delta function \citep{hht2020}. Hence, if we construct an approximation with a representation of order \(2p\) we expect it to perform only marginally better than an approximation constructed with representations of order \(2p-1\). However, the computational cost of the latter is lower, so we opt for the order \(2p-1\) representation.

From the definition of the class of CME distributions, for a given CME with odd order, \(p\), and representation \((\bs\alpha,\bs S)\), the matrix \(\bs S\) has one real eigenvalue, and \(p-1\) complex eigenvalues and all eigenvalues have the same real part \cite{hht2020}.

For CMEs considered here, the vector function \(\bs k(t)=\cfrac{\bs \alpha e^{St}}{\bs \alpha e^{St}\bs e}\) is periodic with period \(\rho = 2\pi/\omega\) where \(\omega=min_i(|\Im(\lambda_i)|)\), \(\lambda_i\) are the eigenvalues of \(\bs{S}\) and \(\Im(z)\) is the imaginary component of a complex number \(z\). We leverage this fact to simplify some numerical integration in Chapter~\ref{sec: construction and modelling}.

@@

\section{Discontinuous Galerkin}
The discontinuous Galerkin (DG) method is a general methodology for numerically approximating solutions to differential equations. The discontinuous Galerkin method is a finite-element method which approximates differential operators by finite dimensional matrices. In this thesis we use the discontinuous Galerkin method to approximate the infinitesimal generator of a fluid queue. Here, we give a very brief introduction to discontinuous Galerkin methods in general and refer the reader to Chapter~\ref{ch:galerkin} for details on the application to fluid queues, and to \cite{nodalDGBook} for more details on the method in general.

Intuitively, the DG method works by first partitioning the spatial domain into cells, on each cell a basis of (typically polynomial) functions is defined, then the differential equation is projected onto the basis of polynomials. Thus, the differential equation is discretised it two ways, by the partition of space into cells, then by the projection onto a finite dimensional basis. Linking the cells together essentially amounts to defining boundary conditions for each cell. To form the boundary condition for cells adjacent to the boundary we take the boundary condition defined by the problem we are solving and discretise it. For cells in the interior the boundary condition is defined by the solution on the adjacent cell. In the context of conservation laws (which is what we solve in this thesis), the boundary conditions on each cell are known as the \emph{flux}, as they describe the rate, with respect to time, that the conserved quantity flows across the boundaries of the cell. The flux is discretised and is known as the \emph{numerical flux}.

Typically, polynomial functions are used as the basis for the DG method. The main choices to make in defining the scheme are the grid which defines the cells, the order of the polynomial basis, and the choice of numerical flux.

\cite{nodalDGBook} claim that the first discontinuous Galerkin finite element scheme was developed in \cite{reed1973} where they use the method to solve the first order differential equation
\[\sigma u + \nabla \cdot (a u) = f,\]
in a 2-dimensional spatial domain, where \(\sigma\) is a constant, \(a\) is a piecewise constant function, and \(u\) is the unknown. The literature has since expanded and includes (but is not limited to) theory on the convergence rate of the scheme, the choice of \emph{numerical flux} used (we will define this later), the choice of spatial discretisation used, and applications to a range of other problems including higher-order differential equations and to partial differential equations (PDEs) such as non-linear conservation law. The PDE which describes the evolution of the distribution of a fluid queue is a conservation law as probability must be conserved.

In one dimension, linear conservation laws take the form 
\begin{align}
	&\cfrac{\partial u( x, t)}{\partial t} - \cfrac{\partial}{\partial x} ( c(t) u( x, t)) = 0,\,  x \in \Omega, \label{eqn: cons law}
	\\& u( x,t)= g( x,t),\tag{boundary condition}
	\\& u( x,0)=f( x).\tag{initial condition}
\end{align}
where \(\Omega\) is the region of interest. Here \(u\) is the conserved quantity, which we will refer to as \emph{mass}, but it could really be any conserved quantity.

To solve conservation laws with the DG method the spatial domain \(\Omega\) is partitioned into cells (intervals) \(\mathcal D_k = [y_k,y_{k+1}]\), \(k=0,...,K-1\). For each cell \(\calD_k\) we choose \(p_k\) linearly independent functions \(\{\phi^r_k\}_{r=1}^{p_k}\), compactly supported on \(\calD_{k}\) (i.e.~\(\phi^r_k(x)=0\) for \(x\notin\calD_{k}\)). Typically, we take \(\phi_k^r(x)\) to be a basis of polynomial functions. The functions \(\{\phi^r_k\}_{r=1}^{p_k}\) span some function space, \(W_k\) say, and further form a basis for this space. It is convenient in this work to take \(\{\phi^r_k\}_{r=1}^{p_k}\) as a basis of \emph{Lagrange interpolating polynomials} defined by the \emph{Gauss-Lobatto quadrature points} \citep{nodalDGBook} (see also Section~\ref{eqn: la}). For the sake of illustration, the reader may think of \(\{\phi^r_k\}_{r=1}^{p_k}\) as the Lagrange polynomials. On each cell \(\mathcal D_{k}\) we approximate 
\[u(x,t)\approx u_{k}(x,t)=\sum\limits_{r=1}^{p_k}a_{k}^r(t)\phi^r_k(x),\] 
where \(a_{k}^r(t)\) are yet-to-be-determined time-dependent coefficients. The approximation \(u_{k}\) is referred to as the \textit{local} approximation on cell \(k\), while the \textit{global} approximation is given by \(\sum\limits_{k=0}^{K-1} u_{k}\) on the whole domain \(\Omega\). The whole approximation space is \(\bigoplus\limits_{k\in\mathbb Z} W_k\), the direct sum of the spaces \(W_k\), \(k\in\{0,...,K-1\}\).

First, consider multiplying (\ref{eqn: cons law}) by the row vector \(\bs \phi_k(x)=[\phi_k^1(x),...,\phi_k^{p_k}(x)]\) and integrating by parts over \(\calD_k\),
\begin{align}
	&\int_{x\in \calD_k}\cfrac{\partial u(x,t)}{\partial t}\bs\phi_k(x)\wrt x + \int_{x\in \calD_k} c(t) u(x,t)\cfrac{\partial}{\partial x} \bs \phi_k( x)\wrt x - c(t)[u(x,t)\bs \phi_k(x)]_{x=y_k}^{y_{k+1}} = 0.\label{eqn: hadndjj}
\end{align}
For a sufficiently large basis of functions, then (\ref{eqn: hadndjj}) is equivalent to the PDE (\ref{eqn: cons law}). The term \(- c(t)[u(x,t)\bs \phi_k(x)]_{x=y_k}^{y_{k+1}}\) is known as the \emph{flux} as it describes the rate at which mass flows into and out of the cell \(\calD_k\). The flux is essentially a boundary condition on the cell \(\calD_k\) and is specified by the value of the solution on the adjoining cells. When \(c(t)\) is positive (negative) then mass flows into (out of) cell \(\calD_k\) at the left-hand boundary and out of (into) cell \(\calD_k\) at the right-hand boundary. Suppose that \(c(t)\) is positive. When \(u\) is continuous, then \(u(y_{k}^-,t)=u(y_k,t),\) where \(y^-\) is the left limit at \(y\), hence the flux into cell \(\calD_k\) in Equation (\ref{eqn: cons law}) is \(c(t)u(y_k^-,t)\bs \phi_k(y_k)\). The flux out of cell \(\calD_k\) is \(c(t)u(y_{k+1}^-,t)\bs \phi_k(y_{k+1})\). Using this, rewrite (\ref{eqn: hadndjj})
\begin{align}
	&\int_{x\in \calD_k}\cfrac{\partial u(x,t)}{\partial t}\bs\phi_k(x)\wrt x + \int_{x\in \calD_k} c(t) u(x,t)\cfrac{\partial}{\partial x} \bs \phi_k( x)\wrt x \nonumber
	\\&\qquad\qquad{}- c(t)[u(y_{k+1}^-,t)\bs \phi_k(y_{k+1}) - u(y_{k}^-,t)\bs \phi_k(y_{k})] = 0.\label{eqn: hadndjjj}
\end{align}
We can proceed analogously when \(c(t)\) is negative.

Upon substituting the approximations \(u_k\) and \(u_{k-1}\) into (\ref{eqn: hadndjjj}) 
\begin{align}
	&\int_{x\in \calD_k}\cfrac{\partial\bs a_k(t)}{\partial t}\bs \phi_k(x)\tr{}\bs\phi_k(x)\wrt x + \int_{x\in \calD_k} c(t) \bs a_k(t)\bs\phi_k(x)\tr{}\cfrac{\partial}{\partial x} \bs \phi_k( x)\wrt x \nonumber
	\\&\qquad\qquad{}- c(t)[\bs a_k(t)\bs\phi_k(y_{k+1}^-)\tr{}\bs \phi_k(y_{k+1}) - \bs a_{k-1}(t)\bs\phi_{k-1}(y_{k}^-)\tr{}\bs \phi_k(y_{k})].\nonumber
	\\&=\cfrac{\wrt \bs a_k(t)}{\wrt t}\bs M_k + c(t) \bs a_k(t)\bs G_k - c(t)[\bs a_k(t)\bs F_{k,k} - \bs a_{k-1}(t)\bs F_{k-1,k}]\nonumber
	\\&= 0,\label{eqn: hadndj}
\end{align}
where 
\[\bs M_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\boldsymbol \phi_k(x)\wrt x,\quad \bs G_k := \int_{x\in\calD_k}\boldsymbol \phi_k(x)\tr{}\cfrac{\wrt}{\wrt x} \boldsymbol \phi_k(x)\wrt x,\]
\[\bs F_{k,k} := \bs\phi_k(y_{k+1}^-)\tr{}\bs \phi_k(y_{k+1}),\quad \bs F_{k-1,k} := \bs\phi_{k-1}(y_{k}^-)\tr{}\bs \phi_k(y_{k}),\]
and are known as the \emph{mass}, \emph{stiffness} and \emph{flux} matrices. Thus, the approximation has reduced the problem to a system of ordinary differential equations. One can determine the coefficients \(\bs a_k(t)\), \(k=0,...,K-1\), from the initial condition \(\bs a_k(0)\), \(k=0,...,K-1\), then integrating over time, perhaps numerically. The initial coefficients \(\bs a_k(0)\), \(k=0,...,K-1\), can be found by projecting the initial condition onto the basis of polynomial functions, \(\{\phi_k^r\}_{k,r}\).

One subtlety is the choice of numerical flux. We arrived at the numerical flux above using a continuity argument. However, the approximate solution \(\sum_{k=0}^{K-1}u_k(x,t)\) is discontinuous at the points \({y_{k}}\). It turns out that the choice of numerical flux is not unique and there are range of values we may choose from. The numerical flux above is known as the \emph{upwind} flux -- we use it throughout this thesis and do not explore others. Other choices which define a discontinuous Galerkin scheme are the basis functions used and the cell geometries. We refer the reader to \citep{nodalDGBook} and reference therein for more details.

Typically, DG schemes are used to solve physical problems in 2 or 3 dimensions. The cell geometries become more complex in higher dimensions and this complicates matters somewhat, but the underlying principles and intuition is the same. In this thesis we solve problems with one spatial dimension only.

Two properties make the DG scheme a natural choice to solve the problem present in this thesis. High-order accuracy enabled by the use of basis functions to model the solution on each cell, and the cell-based structure of the method with cells related to each other at cell edges only and only to neighbouring cells. Finite difference methods can achieve high-order accuracy, however they are not cell-based, which makes them not a natural choice to solve the problems in this thesis. Finite volume methods are cell-based, but do not have the same high-order accuracy as the DG method. Further, approximating flux terms with greater accuracy in the finite volume method can introduce dependencies between cells which are not neighbours and this creates difficulties for the application in this thesis.
